Chapitre 1

### 2023-07-12 Thèse/Chapitre3/Chapitre3B.tex
Rmqs Julie + Article final

\chapter{Valorisation des données hydrométriques anciennes pour l’analyse fréquentielle des crues}
\label{chap:ch3}
\markboth{Valorisation des données hydrométriques anciennes}{Valorisation des données hydrométriques anciennes}

\paragraph{} Ce chapitre prend la forme d'un article scientifique soumis à la revue "Journal of Hydrology" le 18/01/2023 et accepté sous réserve de révisions mineures.

@@

L'analyse fréquentielle des crues est une méthode largement utilisée pour estimer le risque d'inondation. Elle est affectée par plusieurs sources d'incertitude. L'extension des chroniques de débit par l'utilisation de chroniques de hauteurs d'eau anciennes a le potentiel de réduire cette incertitude l'échantillonnage. Les débits anciens estimés sont généralement affectés par de grandes incertitudes. Cet article étudie l'intérêt de l'utilisation de chroniques de hauteurs d'eau anciennes pour améliorer les estimations des quantiles de crue extrêmes à travers une chaine de propagation diverses incertitudes. Les incertitudes sont estimées et propagées à partir des mesures limnimétriques et des courbes de tarages jusqu'aux quantiles de crue à l'aide de procédures Monte Carlo. La part des incertitudes hydrométriques et des incertitudes d'échantillonnage dans les estimations de quantiles de crue est examinée. Cette procédure est appliquée à la série continue de hauteurs d'eau longue de 205 ans à la station hydrométrique du Rhône à Beaucaire, France (95 590 km\textsuperscript{2}). L'incertitude à 95\% des débits maximum annuels estimés varie de 30\% (XIX\textsuperscript{ème} siècle) à 5\% (1967-2020). L'incertitude totale des quantiles de crues est considérablement réduite lorsque la longueur de la série de débits utilisée passe de 20 à 100 ans en raison de la réduction de l'incertitude d'échantillonnage. Cependant, l'incertitude totale reste stable au-delà de cette taille d'échantillon : ceci est dû au fait que d'importantes incertitudes affectent les débits des crues du XIX\textsuperscript{ème} siècle, ce qui compense la réduction de l'incertitude d'échantillonnage. L'utilisation de la chronique totale longue de deux siècles conduit à inclure les deux plus fortes crues connues, en 1840 et 1856. Cela induit une augmentation de 15\% de l'estimation de la crue millénale, ce qui représente une augmentation mineure compte tenu des fortes incertitudes

\selectlanguage{english}
\section{Abstract}

    Flood frequency analysis (FFA), a widely used method to estimate flood hazard, is affected by several sources of uncertainty. Extending flood samples by reanalyzing historical continuous stage records has the potential to reduce sampling uncertainty, but the historical flood discharges derived from this reanalysis are generally affected by large uncertainties. This paper explores whether historical stage records improve design flood estimates through a chain of uncertainty estimation methods for FFA. Uncertainties are estimated and propagated from stage and rating curves to design flood estimates using Monte Carlo procedures. The role of both streamflow and sampling uncertainties in design flood estimation is examined. This procedure  is applied to the 205-year long continuous stage series of the Rhône River at Beaucaire, France (95 590 km\textsuperscript{2}). The estimated streamflow 95\% uncertainty varies from 30\% (XIX\textsuperscript{th} Century) to 5\% (1967-2020). The total uncertainty of design flood is significantly reduced when the length of the series increases from 20 to 100 years due to sampling uncertainty reduction. However, the total uncertainty remains stable beyond this sample size: this is because large uncertainties affecting the XIX\textsuperscript{th} Century flood discharges compensate for the reduction in sampling uncertainty. Enlarging the sample size to two centuries leads to including the two largest known floods in 1840 and 1856. In turn, this induces a 15\% increase of the 1000-year flood estimates, a minor difference considering the associated uncertainty.
    \newline
    \textbf{Keywords}: Flood frequency analysis, Historical stage records, Uncertainty propagation, Streamflow uncertainty, Sampling uncertainty

\section{Introduction}
    \paragraph{}
    Flood frequency analysis (FFA) is a widely used method to estimate flood hazard. It allows linking the magnitude of a flood to its probability of occurrence (\citet{hamed_flood_2019}; \citet{jain_design_2019}). Flood estimates for various exceedance probabilities or, equivalently, return periods, are commonly used for population safety policies, land use planning, as well as industrial safety. The standard FFA approach is to estimate a distribution using a sample of flood peaks, typically defined as annual maximum discharges or discharges over a given threshold. This distribution may be extrapolated to reach the desired flood quantile that typically corresponds to a 100 or 1000-year return period (see \cite{le_delliou_recommandations_2014} for dam safety regulations in France).

    \paragraph{}
    This FFA approach is affected by various sources of uncertainty. First, the hydrological data used to estimate the FFA distribution is uncertain. Indeed, streamflow time series are generally derived from stage time series through rating curve models \citep{rantz_measurement_1982}. This procedure includes measurement (gaugings and stage) and model (rating curve) errors. Moreover, the estimated FFA distribution is also affected by sampling uncertainty resulting from the limited size of available streamflow data \citep{kjeldsen_uncertainty_2011}. Considering the importance of decisions relying on FFA results, a consistent treatment of uncertainty all over the data processing chain (including both streamflow and sampling uncertainties) is essential but is usually not performed.

@@

    \paragraph{}
    Rating curve uncertainty is also a major issue when dealing with streamflow series. Transforming stage into discharge requires calibration data (gaugings) to establish the stage-discharge relationship. Gaugings uncertainty depends on the measurement method (\citet{lecoz_quantification_2014}; \citet{puechberty_charte_2017}). Moreover, the rating curve is also affected by uncertainties coming from the imperfection of the chosen model to represent the actual hydraulic configuration, and from parameter estimation. Many methods have been proposed to quantify these uncertainties (\citet{petersen-overleir_bayesian_2009}; \citet{juston_rating_2014}; \citet{le_coz_combining_2014}; \citet{morlot_dynamic_2014}; \citet{coxon_novel_2015}; \citet{mcmillan_rating_2015}; \citet{mansanarez_rapid_2019}). A comparison of several of these methods has been recently proposed by \citet{kiang_comparison_2018}. Another important issue affecting streamflow data accuracy is rating changes. The stage-discharge relationship is frequently affected by changes caused by various factors, either natural or anthropic, for instance: bed geometry evolution during floods or river works, aquatic vegetation growth and decay, ice cover... A regular monitoring through gaugings is essential to detect those changes \citep{ibbitt_gauging_1987} that can be transient or sudden. Several methods have been proposed to deal with rating changes: estimating rating curves on moving temporal windows (\citet{westerberg_stage-discharge_2011}; \citet{guerrero_temporal_2012}), computing as many rating curves as there are gaugings \citep{morlot_dynamic_2014}, exploring changes in the annual minimum stages \citep{lapuszek_methods_2015}, selecting the 0.5-year return period discharge as a threshold for rating changes \citep{mcmillan_impacts_2010}. More recently, \citet{darienzo_detection_2021} proposed a method based on a recursive segmentation procedure, accounting for both gaugings and rating curve uncertainties. This method has a particular interest when dealing with old and uncertain gaugings. Following the detection of rating shifts, rating curves should be estimated for each stability period. This task may not be straightforward, as the number of gaugings available within a stability period is not always sufficient to properly estimate the stage-discharge relationship for the whole discharge range. A common way to address this problem is to artificially repeat some gaugings from other stability periods (\citet{mcmillan_benchmarking_2012}; \citet{puechberty_charte_2017}). \citet{mansanarez_shift_2019} proposed an alternative approach to deal with this issue. They developed a stage-period-discharge (SPD) model where rating curve parameters may vary across periods, while others are supposed constant. This method has the advantage of transferring information between periods to improve the rating curve estimation, even when few gaugings are available.

    \paragraph{}
    Estimating sampling uncertainty in FFA is a well-established approach. Whatever the chosen distribution and estimation method, standard statistical procedures are available \citep{coles_classical_2001}. However, these standard procedures only quantify sampling uncertainty, they do not consider data uncertainty. The literature review proposed in the previous paragraphs shows that methods for quantifying individual sources of uncertainty (stage, rating curve and FFA distribution estimation) are available. However, the way these multiple uncertainties propagate through the FFA analysis chain has been less thoroughly studied. A few solutions have emerged to propagate uncertainties in stage time series through uncertain rating curves (\citet{dymond_accuracy_1982}; \citet{herschy_hydrometry_1998}; \citet{petersen-overleir_uncertainty_2005}), but they assume independent stage errors and therefore neglect systematic errors. \citet{horner_impact_2018} proposed a method for the propagation of both sources of stage uncertainty through uncertain rating curves. Therefore, it is possible to distinguish the effects of independent and systematic stage errors on streamflow uncertainty. \citet{petersen-overleir_accounting_2009}, \citet{steinbakk_propagation_2016}, and \citet{vieira_assessing_2022} performed an integrated analysis in which both rating curve parameters and flood frequency distribution are estimated. These studies highlighted the importance of considering rating curve uncertainty for design flood estimations and concluded that, under some conditions, accounting for rating curve uncertainty may notably widen the uncertainty intervals around flood quantiles. However, these studies did not consider stage measurement and time interpolation errors nor rating changes, which may constitute a major source of uncertainty for streamflow data. This is particularly the case when dealing with long streamflow series for which stage uncertainty is large and variable through time, and rating changes may have been missed. Their consideration in a flood frequency framework is therefore not avoidable.

@@

    \paragraph{}
    Note that the "historical data" term used in this paper refers to the use of ancient but regularly measured stage series, as opposed to sporadic flood marks, prior to regular stage measurements.

    \paragraph{}
    This paper illustrates the chained application of methods to quantify and propagate uncertainty from stage records (and their limited time resolution) and stage-discharge rating curves to the estimation of design flood. While most of these methods already exist, a key novelty of this work is their combination in a consistent framework (Figure \ref{fig:ChProp}) to provide an end-to-end evaluation of the  uncertainty affecting FFA estimates. An original method to quantify the stage uncertainty stemming from infrequent readings is also proposed.

    The paper is organized as follows. First, the methodology for establishing uncertain streamflow series in a century-long context is presented. It goes through the detection of rating shifts (section \ref{subsec:RatingShifts}), the estimation of rating curves (section \ref{subsec:RC_SPD}), and the estimation (section \ref{subsec:StageErr}) and propagation (section \ref{subsec:PropagStage}) of stage errors. Then, an approach to propagate streamflow uncertainty through the estimation of extreme flood quantiles is proposed (section \ref{subsec:FFA}). This procedure is applied to the Beaucaire gauge on the Rhône River (section \ref{sec:Bcr}), which official design flood estimates is based on a 80-year long discharge series \citep{rigaudiere_etude_2000}. The recent works of \citet{pichard_hydro-climatology_2017} and \citet{bard_actualisation_2018} provided a continuous stage series from 1816 to the present time, which makes it the ideal case study for demonstrating this procedure. The results of this application are presented in section \ref{sec:Results}, and they are discussed in section \ref{sec:Discussion}, where avenues for improvements are proposed.

    \begin{figure}[h!]
    \centering
        \includegraphics[width = 8cm]{Chapitre3/Figures//1-uTotSchema.png}
        \caption{Block diagram of the uncertainty propagation procedure. Grey blocks represent data, blue blocks stand for analysis methods/models that correspond to the sub-sections of this article. $h$ is the water stage, $uh$ the stage uncertainty, $Q$ the discharge, $uQ$ the discharge uncertainty, $t$ is the time and $RP$ the return period of flood quantiles.}
        \label{fig:ChProp}
    \end{figure}
    \FloatBarrier

@@

    \begin{figure}[h!]
        \centering
        \includegraphics[width = 11cm]{Chapitre3/Figures//2-BcrBv.png}
        \caption{The French Rhône River catchment and Beaucaire gauging stations (from www.geoportail.gouv.fr and www.openstreetmap.org)}
        \label{fig:locstations}
    \end{figure}
    % \FloatBarrier

    \subsection{Rating curves}
        \label{sec:prior_elicitation}
        \subsubsection{Pont de Beaucaire}

@@

        \begin{figure}[h!]
            \centering
            \begin{subfigure}{0.7\linewidth}
            \centering
            \includegraphics[width=1\linewidth]{Chapitre3/Figures//3a-DiguesArmand.png}\hfill
            \caption{}
            \label{subfig:diguesarmand}
            \end{subfigure}

            \begin{subfigure}{0.6\linewidth}
            \centering
            \includegraphics[width=1\linewidth]{Chapitre3/Figures//3b-MatrixChenal_EN.png}
            \caption{}
            \label{subfig:matrixCh}
            \end{subfigure}

            \begin{subfigure}{.9\linewidth}
            \includegraphics[width=.48\linewidth]{Chapitre3/Figures//3c-ProfilsBardRestit.png}
            \includegraphics[width=.48\linewidth]{Chapitre3/Figures//3c-ProfilAvalRestitEN.png}
            \caption{}
            \label{subfig:avalprofilesRestit}
            \end{subfigure}

@@

    \begin{figure}[h!]
        \centering
        \includegraphics[width = 15cm]{Chapitre3/Figures//5-Quant5perc_both.png}
        \caption{Time series of annual 5\% stage quantile at both Pont de Beaucaire (1816-1967) and Beaucaire Restitution (1970-2020) stations.}
        \label{fig:quantile5_both}
    \end{figure}
    % \FloatBarrier

    \begin{table}[h!]
        \begin{tabular}{|l|l|l|l|l|}
            \firsthline
            Physical param. & Meaning & Prior & Inferred param. & Prior\\
            \hline
            \multicolumn{5}{|l|}{\textbf{Control 1: main channel}} \\
            $b_1 [m]$      &   Offset              &  $\mathcal{N}(-4,0.5)$   &     $b_1 [m]$    &  $\mathcal{N}(-4,0.5)$ \\
            \hline
            $B_1 [m]$     &   Channel width   &  $\mathcal{LN}(ln(300),0.16)$&$a_1 [m^{3/2}/s]$  & $\mathcal{LN}(ln(128.6),1.8.10^{-2})$\\
            $K_1 [m^{1/3}/s]$&   Strickler coeff. &  $\mathcal{LN}(ln(35),0.14)$    &                &                     \\
            $S_1 [m/m]$     &   Bed slope        &  $\mathcal{LN}(ln(1.5.10^{-4}),0.55)$         &                &  \\
            \hline
            $c_1 [-]$     &   Exponent            &  $\mathcal{N}(5/3,0.025)$&     $c_1 [-]$     &$\mathcal{N}(5/3,0.025)$\\
            \hline
            \multicolumn{5}{|l|}{\textbf{Control 2: floodway}} \\
            %\hline
            $b_2 [m]$     &   Offset              &  $\mathcal{N}(1.5,0.5)$   &     $b_2 [m]$   &  $\mathcal{N}(2,0.5)$ \\
            \hline
            $B_2 [m]$     &   Channel width   &  $\mathcal{LN}(ln(500),0.1)$  &   $a_2 [m^{3/2}/s]$&  $\mathcal{LN}(ln(241.9),1.10^{-2})$\\
            $K_2 [m^{1/3}/s]$&   Strickler coeff. &  $\mathcal{LN}(ln(30),0.16)$    &                 &                     \\
            $S_2 [m/m]$     &   Bed slope        &   $\mathcal{LN}(ln(2.6.10^{-4}),0.34)$        &                &\\
            \hline
            $c_2 [-]$     &   Exponent           &  $\mathcal{N}(5/3,0.025)$&     $c_2 [-]$    &$\mathcal{N}(5/3,0.025)$\\
            \hline
            \multicolumn{5}{|l|}{\textbf{Structural uncertainty parameters}} \\
            %\hline
            $\gamma_{1} [m^{3}/s]$ & Intercept & $\mathcal{U}(0,1000)$ & $\gamma_{1} [m^{3}/s]$ & $\mathcal{U}(0,1000)$\\
            $\gamma_{2} [-]$ & Slope & $\mathcal{U}(0,100)$ & $\gamma_{2} [-]$ &$\mathcal{U}(0,100)$  \\
            \hline
            \multicolumn{5}{|l|}{\textbf{Multiperiod RC parameters}} \\
            $\Delta l [m]$     &   Local change    &  $\mathcal{N}(0,0.3)$&      $\Delta l [m]$     &$\mathcal{N}(0,0.3)$\\
            $\Delta g [m]$     &   overall change       &  $\mathcal{N}(0,0.3)$&      $\Delta g [m]$     &$\mathcal{N}(0,0.3)$\\
            \lasthline
        \end{tabular} 
        \caption{Priors elicitation for Pont de Beaucaire rating curves. $\mathcal{U}(a,b)$ stands for continuous uniform distribution with bounds $a$ and $b$, $\mathcal{N}(\mu,\sigma)$ for Normal distribution with mean $\mu$  and standard deviation $\sigma$, and $\mathcal{LN}(\mu,\sigma)$ for Log Normal distribution with mean $\mu$ and standard deviation $\sigma$.}
        \label{tab:PriorPt}
    \end{table}

>>>>

\chapter{Flood frequency analysis with a 200-year long stage series}
\label{chap:ch3}
%\markboth{Valorisation des données hydrométriques anciennes}{Valorisation des données hydrométriques anciennes}

\begin{figure}[h]
	\centering
		\includegraphics[width=.9\linewidth]{Chapitre3/Figures/Pyrimont.jpg}
        \caption{Jaugeage du Rhône à Pyrimont. Collection CNR}	
\end{figure}

\newpage
\FloatBarrier
\paragraph{} Ce chapitre est constitué d'un article scientifique publié dans la revue "Journal of Hydrology" et accepté le 16/06/2023.

@@

L'analyse fréquentielle des crues est une méthode largement utilisée pour estimer le risque d'inondation. Elle est affectée par plusieurs sources d'incertitude. L'extension des chroniques de débit par l'utilisation de chroniques de hauteurs d'eau anciennes a le potentiel de réduire cette incertitude d'échantillonnage. Les débits anciens estimés sont généralement affectés par de grandes incertitudes. Cet article étudie l'intérêt de l'utilisation de chroniques de hauteurs d'eau anciennes pour améliorer les estimations des quantiles de crue extrêmes à travers une chaine de propagation diverses incertitudes. Les incertitudes sont estimées et propagées à partir des mesures limnimétriques et des courbes de tarages jusqu'aux quantiles de crue à l'aide de procédures Monte Carlo. La part des incertitudes hydrométriques et des incertitudes d'échantillonnage dans les estimations de quantiles de crue est examinée. Cette procédure est appliquée à la série journalière de hauteurs d'eau longue de 205 ans à la station hydrométrique du Rhône à Beaucaire, France (95 590 km\textsuperscript{2}). L'incertitude à 95\% des débits maximum annuels estimés varie de 30\% (XIX\textsuperscript{ème} siècle) à 5\% (1967-2020). L'incertitude totale des quantiles de crues est considérablement réduite lorsque la longueur de la série de débits utilisée passe de 20 à 100 ans en raison de la réduction de l'incertitude d'échantillonnage. Cependant, l'incertitude totale reste stable au-delà de cette taille d'échantillon : ceci est dû au fait que d'importantes incertitudes affectent les débits des crues du XIX\textsuperscript{ème} siècle, ce qui compense la réduction de l'incertitude d'échantillonnage. L'utilisation de la chronique totale longue de deux siècles conduit à inclure les deux plus fortes crues connues, en 1840 et 1856. Cela induit une augmentation de 15\% de l'estimation de la crue millénale, ce qui représente une augmentation mineure compte tenu des fortes incertitudes

\section{Abstract}

    Flood frequency analysis (FFA), a widely used method to estimate flood hazard, is affected by several sources of uncertainty. Extending flood samples by reanalyzing historical systematic stage records has the potential to reduce sampling uncertainty, but the historical flood discharges derived from this reanalysis are generally affected by large uncertainties. This paper explores whether historical stage records improve design flood estimates through a chain of uncertainty estimation methods for FFA. Uncertainties are estimated and propagated from stage and rating curves to design flood estimates using Monte Carlo procedures. The role of both streamflow and sampling uncertainties in design flood estimation is examined. This procedure  is applied to the 205-year long systematic stage series of the Rhône River at Beaucaire, France (95 590 km\textsuperscript{2}). The estimated streamflow 95\% uncertainty varies from 30\% (XIX\textsuperscript{th} Century) to 5\% (1967-2020). The total uncertainty of design flood is significantly reduced when the length of the series increases from 20 to 100 years due to sampling uncertainty reduction. However, the total uncertainty remains stable beyond this sample size: this is because large uncertainties affecting the XIX\textsuperscript{th} Century flood discharges compensate for the reduction in sampling uncertainty. Enlarging the sample size to two centuries leads to including the two largest known floods in 1840 and 1856. In turn, this induces a 15\% increase of the 1000-year flood estimates.
    \newline
    \textbf{Keywords}: Flood frequency analysis, Historical stage records, Uncertainty propagation, Streamflow uncertainty, Sampling uncertainty

\section{Introduction}
    \paragraph{}
    Flood frequency analysis (FFA) is a widely used method to estimate flood hazard. It allows linking the magnitude of a flood to its probability of occurrence (\citet{hamed_flood_2019}; \citet{jain_design_2019}). Flood estimates for various exceedance probabilities or, equivalently, return periods, are commonly used for population safety policies, land use planning, as well as industrial safety. The standard FFA approach is to estimate a distribution using a sample of flood peaks, typically defined as annual maximum discharges or discharges over a given threshold. This estimated distribution may be extrapolated to reach the desired flood quantile that typically corresponds to a 100 or 1000-year return period (see \cite{le_delliou_recommandations_2014} for dam safety regulations in France).

    \paragraph{}
    This FFA approach is affected by various sources of uncertainty. First, the hydrological data used to estimate the FFA distribution is uncertain. Indeed, streamflow time series are generally derived from stage time series through rating curve models \citep{rantz_measurement_1982}. This procedure includes three types of errors on hydrological data: stage measurement; discharge measurement (gaugings); stage-discharge model (rating curve). Moreover, the estimated FFA distribution is also affected by sampling uncertainty resulting from the limited size of available streamflow data \citep{kjeldsen_uncertainty_2011}. Considering the importance of decisions relying on FFA results, a consistent treatment of uncertainty all over the data processing chain (including both streamflow and sampling uncertainties) is essential but is usually not performed.

@@

    \paragraph{}
    Rating curve uncertainty is also a major issue when dealing with streamflow series. Transforming stage into discharge requires calibration data (gaugings) to establish the stage-discharge relationship. Note that the term "gauging" corresponds in this paper (and in the French practice) to the sporadic measurement of the stage-discharge couple, which can correspond to the term "discharge measurement" in the British or American practice. Gaugings uncertainty depends on the measurement method (\citet{lecoz_quantification_2014}; \citet{puechberty_charte_2017}). Moreover, the rating curve is also affected by uncertainties coming from the imperfection of the chosen model to represent the actual hydraulic configuration, and from parameter estimation. Many methods have been proposed to quantify these uncertainties (\citet{petersen-overleir_bayesian_2009}; \citet{juston_rating_2014}; \citet{le_coz_combining_2014}; \citet{morlot_dynamic_2014}; \citet{coxon_novel_2015}; \citet{mcmillan_rating_2015}; \citet{mansanarez_rapid_2019}). A comparison of several of these methods has been recently proposed by \citet{kiang_comparison_2018}. Another important issue affecting streamflow data accuracy is rating changes. The stage-discharge relationship is frequently affected by changes caused by various factors, either natural or anthropic, for instance: bed geometry evolution during floods or river works, aquatic vegetation growth and decay, ice cover... A regular monitoring through gaugings is essential to detect those changes \citep{ibbitt_gauging_1987} that can be transient or sudden. Several methods have been proposed to deal with rating changes: estimating rating curves on moving temporal windows (\citet{westerberg_stage-discharge_2011}; \citet{guerrero_temporal_2012}), computing as many rating curves as there are gaugings \citep{morlot_dynamic_2014}, exploring changes in the annual minimum stages \citep{lapuszek_methods_2015}, selecting the 0.5-year return period discharge as a threshold for rating changes \citep{mcmillan_impacts_2010}. More recently, \citet{darienzo_detection_2021} proposed a method based on a recursive segmentation procedure, accounting for both gaugings and rating curve uncertainties. This method has a particular interest when dealing with old and uncertain gaugings. Following the detection of rating shifts, rating curves should be estimated for each stable period. This task may not be straightforward, as the number of gaugings available within a stable period is not always sufficient to properly estimate the stage-discharge relationship for the whole discharge range. A common way to address the lack of gaugings (in particular flood gaugings) within some of the stable periods is to use gaugings from the other stable periods (\cite{mcmillan_benchmarking_2012}; \cite{puechberty_charte_2017}). \citet{mansanarez_shift_2019} proposed an alternative approach to deal with this issue. They developed a stage-period-discharge (SPD) model where rating curve parameters may vary across periods, while others are supposed constant. This method has the advantage of transferring information between periods to improve the rating curve estimation, even when few gaugings are available.

    \paragraph{}
    Estimating sampling uncertainty in FFA is a well-established approach. Whatever the chosen distribution and estimation method, standard statistical procedures are available \citep{coles_classical_2001}. However, these standard procedures only quantify sampling uncertainty, they do not consider data uncertainty. The literature review proposed in the previous paragraphs shows that methods for quantifying individual sources of uncertainty (stage, rating curve and FFA distribution estimation) are available. However, the way these multiple uncertainties propagate through the FFA analysis chain has been less thoroughly studied. A few solutions have emerged to propagate uncertainties in stage time series through uncertain rating curves (\citet{dymond_accuracy_1982}; \citet{herschy_hydrometry_1998}; \citet{petersen-overleir_uncertainty_2005}), but they assume independent stage errors and therefore neglect systematic errors. \citet{horner_impact_2018} proposed a method for the propagation of both sources of stage uncertainty through uncertain rating curves. Therefore, it is possible to distinguish the effects of independent and systematic stage errors on streamflow uncertainty. \citet{petersen-overleir_accounting_2009}, \citet{steinbakk_propagation_2016}, and \citet{vieira_assessing_2022} performed an integrated analysis in which both rating curve parameters and flood frequency distribution are estimated. These studies highlighted the importance of considering rating curve uncertainty for design flood estimations and concluded that, under some conditions, accounting for rating curve uncertainty may notably widen the uncertainty intervals around flood quantiles. However, these studies did not consider stage measurement and time interpolation errors nor rating changes, which may constitute a major source of uncertainty for streamflow data. This is particularly the case when dealing with long streamflow series for which stage uncertainty is large and variable through time, and rating changes may have been missed. Their consideration of all these sources of uncertainty is necessary.

@@

    \paragraph{}
    Note that the "historical data" term used in this paper refers to the use of old but continuously measured stage series, as opposed to sporadic flood marks, prior to regular stage measurements.

    \paragraph{}
    This paper illustrates the chained application of methods to quantify and propagate uncertainty from stage records (and their limited time resolution) and stage-discharge rating curves to the estimation of flood distribution with uncertainties. While most of these methods already exist, a key novelty of this work is their combination in a consistent framework (Figure \ref{fig:ChProp}) to provide an end-to-end evaluation of the  uncertainty affecting FFA estimates. An original method to quantify the stage uncertainty stemming from infrequent readings is also proposed.

    The paper is organized as follows. First, the methodology for establishing uncertain streamflow series in a century-long context is presented. It goes through the detection of rating shifts (section \ref{subsec:RatingShifts}), the estimation of rating curves (section \ref{subsec:RC_SPD}), and the estimation (section \ref{subsec:StageErr}) and propagation (section \ref{subsec:PropagStage}) of stage errors. Then, an approach to propagate streamflow uncertainty through the estimation of extreme flood quantiles is proposed (section \ref{subsec:FFA}). This procedure is applied to the Beaucaire gauge on the Rhône River (section \ref{sec:Bcr}), which previous official FFA only used a 80-year long discharge series \citep{rigaudiere_etude_2000}. In France, the official design flood for flood risk mapping is based on the “largest known flood” or the Q100 flood (AEP=0.01) if the latter is greater. Uncertainty is not taken into account in the official rules. The recent works of \citet{pichard_hydro-climatology_2017} and \citet{bard_actualisation_2018} provided a systematic stage series from 1816 to the present time, which makes it the ideal case study for demonstrating this procedure. The results of this application are presented in section \ref{sec:Results}, and they are discussed in section \ref{sec:Discussion}, where avenues for improvements are proposed.

    \begin{figure}[h!]
    \centering
        \includegraphics[width = 8cm]{Chapitre3/Figures/1-uTotSchema.png}
        \caption{Block diagram of the uncertainty propagation procedure. Grey blocks represent data, blue blocks stand for analysis methods/models that correspond to the sub-sections of this article. $h$ is the water stage, $uh$ the stage uncertainty, $Q$ the discharge, $uQ$ the discharge uncertainty, $t$ is the time and $RP$ the return period of flood quantiles.}
        \label{fig:ChProp}
    \end{figure}
    \FloatBarrier

@@

    \begin{figure}[h!]
        \centering
        \includegraphics[width = 11cm]{Chapitre3/Figures/2-BcrBv.png}
        \caption{The French Rhône River catchment and Beaucaire gauging stations (from www.geoportail.gouv.fr and www.openstreetmap.org)}
        \label{fig:locstations}
    \end{figure}
    % \FloatBarrier

    \subsection{Rating curves}
    \label{sec:prior_elicitation}

	\paragraph{} Many gauges of the Rhône River are subject to the effect of variable backwaters caused by the proximity of a dam, and therefore require the use of a stage-fall-discharge (SFD) rating curve model (for example, Valence gauge, 140 km upstream from Beaucaire described by \cite{mansanarez_non-unique_2016}). Beaucaire is located within a narrowing of the floodplain and there is no dam downstream from the gauge. However, a backwater effect from the sea has been observed at Beaucaire Restitution, but it only affects the very low flows. As this article focuses on floods, we assume that there is no reason to use a SFD model here. Consequently, like the gauge operator (CNR), a stage-discharge (SD) model is used for both Pont de Beaucaire and Beaucaire Restitution gauges.

        \subsubsection{Pont de Beaucaire}

@@

        \begin{figure}[h!]
            \centering
            \begin{subfigure}{0.7\linewidth}
            \centering
            \includegraphics[width=1\linewidth]{Chapitre3/Figures/3a-DiguesArmand.png}\hfill
            \caption{}
            \label{subfig:diguesarmand}
            \end{subfigure}

            \begin{subfigure}{0.6\linewidth}
            \centering
            \includegraphics[width=1\linewidth]{Chapitre3/Figures/3b-MatrixChenal_EN.png}
            \caption{}
            \label{subfig:matrixCh}
            \end{subfigure}

            \begin{subfigure}{.9\linewidth}
            \includegraphics[width=.48\linewidth]{Chapitre3/Figures/3c-ProfilsBardRestit.png}
            \includegraphics[width=.48\linewidth]{Chapitre3/Figures/3c-ProfilAvalRestitEN.png}
            \caption{}
            \label{subfig:avalprofilesRestit}
            \end{subfigure}

@@

    \begin{figure}[h!]
        \centering
        \includegraphics[width = 15cm]{Chapitre3/Figures/5-Quant5perc_both.png}
        \caption{Time series of annual 5\% stage quantile at both Pont de Beaucaire (1816-1967) and Beaucaire Restitution (1970-2020) stations.}
        \label{fig:quantile5_both}
    \end{figure}
    % \FloatBarrier

    \begin{table}[h!]
        \begin{tabular}{|l|l|l|l|l|}
            \firsthline
            Physical param. & Meaning & Prior & Inferred param. & Prior\\
            \hline
            \multicolumn{5}{|l|}{\textbf{Control 1: main channel}} \\
            $b_1 [m]$      &   Offset              &  $\mathcal{N}(-4,0.5)$   &     $b_1 [m]$    &  $\mathcal{N}(-4,0.5)$ \\
            \hline
            $B_1 [m]$     &   Channel width   &  $\mathcal{LN}(ln(300),0.16)$&$a_1 [m^{3/2}/s]$  & $\mathcal{LN}(ln(128.6),1.8.10^{-2})$\\
            $K_1 [m^{1/3}/s]$&   Strickler coeff. &  $\mathcal{LN}(ln(35),0.14)$    &                &                     \\
            $S_1 [m/m]$     &   Bed slope        &  $\mathcal{LN}(ln(1.5.10^{-4}),0.55)$         &                &  \\
            \hline
            $c_1 [-]$     &   Exponent            &  $\mathcal{N}(5/3,0.025)$&     $c_1 [-]$     &$\mathcal{N}(5/3,0.025)$\\
            \hline
            \multicolumn{5}{|l|}{\textbf{Control 2: floodway}} \\
            %\hline
            $b_2 [m]$     &   Offset              &  $\mathcal{N}(1.5,0.5)$   &     $b_2 [m]$   &  $\mathcal{N}(2,0.5)$ \\
            \hline
            $B_2 [m]$     &   Channel width   &  $\mathcal{LN}(ln(500),0.1)$  &   $a_2 [m^{3/2}/s]$&  $\mathcal{LN}(ln(241.9),1.10^{-2})$\\
            $K_2 [m^{1/3}/s]$&   Strickler coeff. &  $\mathcal{LN}(ln(30),0.16)$    &                 &                     \\
            $S_2 [m/m]$     &   Bed slope        &   $\mathcal{LN}(ln(2.6.10^{-4}),0.34)$        &                &\\
            \hline
            $c_2 [-]$     &   Exponent           &  $\mathcal{N}(5/3,0.025)$&     $c_2 [-]$    &$\mathcal{N}(5/3,0.025)$\\
            \hline
            \multicolumn{5}{|l|}{\textbf{Structural uncertainty parameters}} \\
            %\hline
            $\gamma_{1} [m^{3}/s]$ & Intercept & $\mathcal{U}(0,1000)$ & $\gamma_{1} [m^{3}/s]$ & $\mathcal{U}(0,1000)$\\
            $\gamma_{2} [-]$ & Slope & $\mathcal{U}(0,100)$ & $\gamma_{2} [-]$ &$\mathcal{U}(0,100)$  \\
            \hline
            \multicolumn{5}{|l|}{\textbf{Multiperiod RC parameters}} \\
            $\Delta l [m]$     &   Local change    &  $\mathcal{N}(0,0.3)$&      $\Delta l [m]$     &$\mathcal{N}(0,0.3)$\\
            $\Delta g [m]$     &   overall change       &  $\mathcal{N}(0,0.3)$&      $\Delta g [m]$     &$\mathcal{N}(0,0.3)$\\
            \lasthline
        \end{tabular} 
        \caption{Priors elicitation for Pont de Beaucaire rating curves. $\mathcal{U}(a,b)$ stands for continuous uniform distribution with bounds $a$ and $b$, $\mathcal{N}(\mu,\sigma)$ for Normal distribution with mean $\mu$  and standard deviation $\sigma$, and $\mathcal{LN}(\mu,\sigma)$ for Log Normal distribution with log-mean $\mu$ and log-standard-deviation $\sigma$.}
        \label{tab:PriorPt}
    \end{table}
